---
title: "Analysis of Jeopardy Games"
author: "Stephen Kaluzny"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  ioslides_presentation:
    css: mystyles.css
---

```{R setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
# options(width = 60)
fht <- 12
fwh <- 12
```

```{r echo=FALSE}
jeopardyDataRaw <- readRDS("jeopardyDataRaw.rds")
source("jcleanData.R")
# jeopardyData <- readRDS("jeopardyData.rds")
```

## Jeopardy - The Game
* Television quiz show where three contestants compete to answer quesions
* Their responses must be in the form of a question
* The daily syndicated version of the show has been on the air since 1984.

## Game Description
* The show has two rounds and a final.
* Each round has 6 categories each with 5 clues
* Clues have increasing dollar amounts within a category.

## Final Jeopardy
* Players are told the category
* They can wager as much of their score
* 30 seconds to write down answer
* Player with the highest score wins that amount in dollars
    * plays in the next day's game against two new players.

## Jeopardy Data
* Data scraped from Internet site
* Information about players
    * name, home city, occupation
    * number of correct and wrong answers
    * daily double play
    * number of games played, winnings

## Purpose of This Analysis

This document is an example of a data analysis
using Open Source R.

This is a reproducible analysis.
All code to recreate the analysis, from getting the data,
cleaning it, computing summaries, to creating graphics is included.
This means that the analysis can be rerun at a future time after more Jeopardy games have been played
and the results of those games are posted to the Internet.
The code for is available on Github at
[https://github.com/spkal/jeopardy](https://github.com/spkal/jeopardy).

## Use of TidyVerse

This is also an exercise in using the `tidyverse` packages,
especially the `dplyr` package,
for data cleanup and manipulation.
The `ggplot2` package is used for creating graphics.
We make extensive use of the new pipe operator (`%>%`)
that was introduced wtih the `dplyr` package.

Those key packages are loaded here.
```{r}
library(ggplot2, quietly= TRUE, warn.conflicts = FALSE)
library(dplyr, quietly= TRUE, warn.conflicts = FALSE)
```

## The Data
* The data was scraped from the
[http://www.j-archive.com site](http://www.j-archive.com) site
* Only regular Jeopardy games were downloaded,
leaving out championships, teen, college, etc tournaments.

```{r eval=FALSE}
jeopardyDataRaw <- jscrapeDirect()
```

## Scraped Data
The scraped data has 15 columns:
```{r}
names(jeopardyDataRaw)
```

## Cleaning the Data
* Variables were renamed to a consistent and descriptive naming scheme.
    * values in the wrong column
    * spelling error in names of cities and states,
    * now living in ..." and "now stationed at ..." to listed location
* The cleaning was an iterative process
    * looking at summaries of the data
    * finding more errors,
    * making corrections

## Clean Function
* The clean up operations were all encapsulate in a single function `jcleanData`
* Run on the scraped data

```{r}
jeopardyData <- jcleanData(jeopardyDataRaw)
```

## Additional Variables
* `PlayerId` was created by pasting together the `Name`, `Occupation`, `City`, and `State`:
* Analysis of `PlayerId` indicated that it uniquely identified players
```{r}
jeopardyData$PlayerId <- with(jeopardyData,
    paste(Name, Occupation, City, State, sep='.'))
```
* `Date` variable was extracted from information in the `Title` variable
```{r}
airedIndex <- regexpr(" aired ", jeopardyData$Title)
jeopardyData$Date <- as.Date(
    substr(jeopardyData$Title, airedIndex + 7, nchar(jeopardyData$Title)))
```

## Winning Amount
* After November 23, 2001 the game doubled the dollar amounts for each answer.
* A `GameMaxValue` variable was added to distinguish games before and after the change:

```{r}
jeopardyData$GameMaxValue <- ifelse(jeopardyData$Date < as.Date("2001-11-23"),
    "Max1000", "Max2000")
```

## The Winner
A variable, `IsWinner`, was added to identify the winner(s) in each game
(tied final scores result in multiple winners):

```{r}
isWinner <-
function (x)
{
    r <- rank(x)
    if (all(r == 2)) {
        r <- c(4, 4, 4)
    }
    (r == 3) | (r == 2.5) | (r == 4)
}
jeopardyData <- group_by(jeopardyData, Show) %>%
    mutate(IsWinner = isWinner(FinalScore))
jeopardyData <- ungroup(jeopardyData)
```

## Gender
* Gender of each player estimated using the `gender` package from CRAN
* `gender::gender` function returns the proportion of males and females with a name
* A cutoff proportion of `0.75` used
    * if the proportion for a gender was at least `0.75`, use that gender.
    * if it was less than `0.75`, set to `NA`

## Gender Computation (non-Tidyverse)
```{r eval=FALSE}
library(gender)
firstName <- with(jeopardyData, gsub("^(.*?)\\s.*", "\\1", Name))
uFirstName <- unique(firstName)  # Only do the unique firstNames
uGenderDF <- gender(uFirstName)
uGender <- uGenderDF$gender
uFirstName <- uGenderDF$name
# Code "male" as "M", "female" as "F"
uGender <- ifelse(uGender == "male", "M", uGender)
uGender <- ifelse(uGender == "female", "F", uGender)
# Only consider gender that has prop. greater than propCutoff, otherwise NA
propCutoff <- 0.75
uGender <- ifelse((!is.na(uGender) & uGender == "M" &
    uGenderDF$proportion_male < propCutoff), NA, uGender)
uGender <- ifelse((!is.na(uGender) & uGender == "F" &
    uGenderDF$proportion_female < propCutoff), NA, uGender)
mFirstName <- match(firstName, uFirstName)
jeopardyData$Gender <- uGender[mFirstName]
```

## Gender Computation
```{r gender}
library(gender)
jeopardyData <- jeopardyData %>%
  mutate(firstName = gsub("^(.*?)\\s.*", "\\1", Name))
# Only do the unique firstNames
uGenderDF <- gender(unique(jeopardyData[["firstName"]]))
# Only consider gender that has prop. greater than propCutoff, otherwise NA
propCutoff <- 0.75
uGenderDF <- uGenderDF %>%
  mutate(Gender = case_when(
    proportion_male > propCutoff ~ "M",
    proportion_female > propCutoff ~ "F",
    TRUE ~ NA_character_))
jeopardyData <- jeopardyData %>%
  left_join(uGenderDF[, c("name", "Gender")],
  by=c("firstName" = "name")) %>%
  select(-firstName)
```

## Gender Results
* Gender predictions for
`r with(jeopardyData, sum(!duplicated(PlayerId) & !is.na(jeopardyData$Gender)))`
players
* `r round(with(jeopardyData, sum(!duplicated(PlayerId) & !is.na(jeopardyData$Gender)) / sum(!duplicated(PlayerId))) *100, 2)` percent of the players.

## Final DataSet
* The final dataset had `r ncol(jeopardyData)` variables:
```{r}
names(jeopardyData)
```
* Each row is data for a single player in a game as the basic observation
* Each games consists of 3 rows.
* `r nrow(jeopardyData)` rows and `r ncol(jeopardyData)` columns
* `r nrow(jeopardyData)` player observations and `r nrow(jeopardyData) / 3` games.
* There are `r length(unique(jeopardyData$PlayerId))` unique players.
* Games played from `r min(jeopardyData$Date)` to `r max(jeopardyData$Date)`

## Answers
* 6 categories x 5 clues per category x 2 rounds = 60 clues in each game.
* A single player answers are:
    * the sum of the number correct (`NumRight`)
    * the number wrong (`NumWrong`),
    * number of daily doubles correct (`DDRight`)
    * the number of daily doubles wrong (`DDWrong`)
```{r}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight) %>%
    with(summary(NumAttempted))
```

At least one player only attempted 2 clues and while another dominated
a game by trying 51.

## A graphical view of the number of attempted clues:
```{r plot01code, eval=FALSE}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight) %>%
    ggplot(aes(x=factor(NumAttempted))) +
        geom_bar() + xlab("Number Clues (out of 60 total) Attempted by a Player") +
        ggtitle("Distribution of Number of Clues Attempted by a Player")
```
-----

```{r plot01, fig.show='hold', fig.align='center', echo=FALSE}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight) %>%
    ggplot(aes(x=factor(NumAttempted))) +
        geom_bar() + xlab("Number Clues Attempted (out of 60 total) by a Player") +
        ggtitle("Distribution of Number of Clues Attempted by a Player")
```

## The same graph, by gender

```{r plot02code, eval=FALSE, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight) %>%
    filter(!is.na(Gender)) %>%
    ggplot(aes(x=factor(NumAttempted))) +
        geom_bar() + xlab("Number Clues Attempted by a Player") +
        facet_grid(Gender ~ .) +
        ggtitle("Distribution of Clues Attempted by Gender")

```

-----

```{r plot02, echo=FALSE, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight) %>%
    filter(!is.na(Gender)) %>%
    ggplot(aes(x=factor(NumAttempted))) +
        geom_bar() + xlab("Number Clues Attempted by a Player") +
        facet_grid(Gender ~ .) +
        ggtitle("Distribution of Clues Attempted by Gender")

```

Men tend to attempt more questions then women
and the range of values is wider for men including
some that attempt over 50 of the possible questions.

Note that the sum of `NumAttempted` across all three players in a game
is not the same as the total number of clues tried in that game
since after an incorrect response by a players,
another play can try to answer that same question.

## The percentage of clues answers correctly
```{r}
jeopardyData %>%
    mutate(PercentCorrect =
        (NumRight + DDRight) / (NumRight + NumWrong + DDWrong + DDRight) * 100) %>%
    with(summary(PercentCorrect))
```

## A density plot of the percent correct answers:
```{r plot03, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
jeopardyData %>%
    mutate(PercentCorrect =
        (NumRight + DDRight) /
            (NumRight + NumWrong + DDWrong + DDRight) * 100) %>%
    ggplot(aes(x=PercentCorrect)) +
        geom_density()
```

The most common percent correct values are between 82 and 90,
then it drops off but rises again at 100 percent.

## Always Right
Here is the distribution of attempted clues for those who always
answer correctly compared to other players. We use a density plot
instead of counts since the total numbers in each group are vastly
different.

##

```{r plot04, fig.width=12, fig.height=12, fig.show='hold', fig.align='center'}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight,
        PercentCorrect = (NumRight + DDRight) /
            (NumRight + NumWrong + DDWrong + DDRight) * 100,
        AlwaysRight = (PercentCorrect > 99)) %>%
    ggplot(aes(x = NumAttempted, color=AlwaysRight)) +
        geom_density() +
        xlab("Number of Clues Attempted") +
        ggtitle("Number Attempted for Perfect Player vs Other Players")
```

Most players who get all questions correct tend not to attempt as many questions as other players.

## Density of percent correct answers by gender
```{r plot05, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
jeopardyData %>%
    filter(!is.na(Gender)) %>%
    mutate(PercentCorrect =
        (NumRight + DDRight) /
            (NumRight + NumWrong + DDWrong + DDRight) * 100) %>%
    ggplot(aes(x=PercentCorrect, color=Gender)) +
        geom_density() +
        ggtitle("Density of Percent Correct by Gender")
```

We have the spike at 100 percent for both female and male players.
The female players have a more spread out distribution of percent correct
and their peak is slightly less than the peak for the male players.

## Daily Doubles
There are 3 daily doubles in each game, one in the Jeopardy Round and
two in the Double Jeopardy round.
The player who gets the daily double questions
can wager as much of their current score on the question.

The number of daily doubles found by players in a game:

```{r}
jeopardyData %>%
    mutate(DDCount = DDWrong + DDRight) %>%
    with(cbind(NumDD = 0:3, DDCount = table(DDCount),
       Percent =  round(table(DDCount) /  sum(table(DDCount)) * 100, 2)))
```

## The percentage of all daily doubles answered correctly
`r
jeopardyData %>%
    filter(DDWrong + DDRight > 0) %>%
    summarise(PercentDDRight = round(sum(DDRight) / sum(DDWrong + DDRight) * 100, 2))
` percent.

## Number of Final Jeopardy Players

If a player has a score of 0 or less at the end of the double jeopardy round,
that player is eliminated from the final jeopardy round.
The number of players in Final Jeopardy:

```{r}
NFinal <- jeopardyData %>%
    group_by(Show) %>%
    mutate(NumberInFinal = sum(SecondRoundScore > 0)) %>%
    summarise(NumberInFinal = first(NumberInFinal)) %>%
    with(table(NumberInFinal))
cbind(NumberOfPlayers=1:3, Count = NFinal,
    Percentage = round(NFinal / sum(NFinal) * 100, 2))
```

The shows with only one player left for Final Jeopardy:
```{r}
jeopardyData %>%
    group_by(Show) %>%
    mutate(NumberInFinal = sum(SecondRoundScore > 0)) %>%
    filter(NumberInFinal == 1) %>%
    select(Date, Name, SecondRoundScore) %>%
    as.data.frame()
```

## Runaways
A runaway game is when one player cannot be beat in Final Jeopardy.
This will occur when one player has at least twice the score of each of the
other players at the end of the Double Jeopardy round.

```{r}
isRunaway <- function(score) {
    score <- sort(score)
    all(score[3] >= 2*score[-3])
}
jeopardyData %>%
    group_by(Show) %>%
    summarise(Runaway = isRunaway(SecondRoundScore)) %>%
    with(c(NumberRunaways=sum(Runaway),
        PercentRunaways=round(sum(Runaway) / length(Runaway) * 100, 2)))
```

## Ties
How often does the game end with multiple winners (i.e. a tie)?

```{r}
jeopardyData %>%
    group_by(Show) %>%
    mutate(NumWinners = sum(IsWinner)) %>%
    summarise(NumWinners = first(NumWinners)) %>%
    with(cbind(Winners = 1:3, Count = table(NumWinners),
        Percent = round(table(NumWinners) / sum(NumWinners) * 100, 2)))
```

Most games have only one winner.
In two games, all three people had the same final score.

## Comebacks

How often does the player with the lowest score
at the end of the double jeopardy round, win the game?

```{r}
jeopardyData %>%
    group_by(Show) %>%
    mutate(IsComeback = (SecondRoundScore == min(SecondRoundScore))) %>%
    filter(IsWinner) %>%
    with(c(Count = sum(IsComeback),
        Percent = round(sum(IsComeback) / length(IsComeback) * 100, 2)))
```

## Final Jeopardy Stumpers

How often do all the players miss the Final Jeopardy questions?
The data does not include any specific information about the Final Jeopardy question
or wager.
If the `FinalScore` is larger than the `SecondRoundScore`, we know the player answered
Final Jeopardy correctly. Similarly, if `FinalScore` is less than `SecondRoundScore`,
their response for Final Jeopardy was incorrect.
If the `FinalScore` is the same as the `SecondRoundScore`, we only know that the
player did not wager anything in Final Jeopardy, their response could have been correct or wrong.
So we only count those who made a wager in Final Jeopardy.

```{r}
isStumper <- function(score2, score3) {
    all(score3 < score2)
}
jeopardyData %>%
    group_by(Show) %>%
    filter(all(SecondRoundScore != FinalScore)) %>%
    summarise(IsStumper = isStumper(SecondRoundScore, FinalScore)) %>%
    with(c(Count = sum(IsStumper), Games = length(IsStumper),
        Percent = round(sum(IsStumper) / length(IsStumper) * 100, 2)))
```

## Dollar Winnings in a Game

When looking at the dollar winnings for a game we need to look separately
at games before the dollar values doubled and after.

```{r}
jeopardyData %>%
    filter(IsWinner) %>%
    group_by(GameMaxValue) %>%
    summarize(AveWinnings=mean(FinalScore),
        MedianWinnings=median(FinalScore),
        MinWinnings=min(FinalScore),
        MaxWinnings=max(FinalScore)) %>%
    as.data.frame()
```

The mean and median winnings are both about doubled
when the dollar amounts doubled.

The shows where the winning score was 0:

```{r}
jeopardyData %>%
    filter(IsWinner, FinalScore == 0) %>%
    select(Date, Name, FinalScore) %>%
    as.data.frame()
```

In one game, all 3 players won with 0
and in another game, two players tied with a score of 0.
Actually, none of the players were considered winners
since none of these players played in the next day's game.

## Distribution of Winnings
```{r plot06, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
filter(jeopardyData, IsWinner) %>%
    ggplot(aes(x=FinalScore)) +
    geom_histogram(binwidth = 2000) +
    facet_grid(GameMaxValue ~ .) +
    ggtitle("Distribution of Game Winnings")
```

There are
`r filter(jeopardyData, GameMaxValue == "Max2000") %>% nrow()`
with GameMaxValue of Max2000
compared to only
`r filter(jeopardyData, GameMaxValue == "Max1000") %>% nrow()`
for Max1000.
It is no surprising that the spread of the FinalScore values
is larger for the Max2000 group.

## The Players
The `NumTimesOnShow` variable increases each time the player is on the show.
We want to look at the maximum value of `NumTimeOnShow` for each player:

```{r}
jeopardyData %>%
    group_by(PlayerId) %>%
    mutate(MaxNumGames = max(NumTimesOnShow)) %>%
    with(summary(MaxNumGames))
```

Half the players only are on one show.
Which is not surprising since two players (usually) do not return.
One player was on 75 times.
We will investigate that player later.


## Games by Gender

Number of games played by gender:

```{r}
jeopardyData %>%
    filter(!is.na(Gender)) %>%
    group_by(PlayerId) %>%
    mutate(MaxNumGames = max(NumTimesOnShow)) %>%
    group_by(Gender) %>%
    summarise(Min=min(MaxNumGames), Median=median(MaxNumGames),
        Mean=mean(MaxNumGames), Max=max(MaxNumGames)) %>%
	as.data.frame()
```

## Most Wins

Looking at the players who won the most games (10 or more),
the Date is the date of their final game:

```{r}
jeopardyData %>%
    group_by(PlayerId) %>%
    summarise(Name = first(Name), MaxNumGames = max(NumTimesOnShow),
        Date = max(Date)) %>%
    filter(MaxNumGames >= 10) %>%
    select(Name, MaxNumGames, Date) %>%
    arrange(desc(MaxNumGames), desc(Date)) %>%
	as.data.frame()
```

##

The distribution of the number of games by a player,
leaving out the outliers (players who played more than 10 games)
as well as players who never won (i.e. only played one game):

```{r plot07, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
jeopardyData %>%
    group_by(PlayerId) %>%
    summarise(MaxNumGames = max(NumTimesOnShow)) %>%
    filter(MaxNumGames > 1 & MaxNumGames < 11) %>%
    ggplot(aes(x=factor(MaxNumGames))) +
        # geom_bar(binwidth=0.10, stat="count") + xlab("NumGamesPlayed") +
        geom_bar() + xlab("NumGamesPlayed") +
        coord_flip() +
        ggtitle("Counts of Number of Games Played by a Player")
```

Most winners only win once and winning more than 5 games is rare.

## Total Winnings Per Player
Here we look at the total winnings for player over all the
games she or he played.
The number of wins is one less than the number of times on the show.

Here are the winnings for the players who won 10 or more games:

```{r}
jeopardyData %>%
    group_by(PlayerId) %>%
    summarise(Name = first(Name), MaxNumGames = max(NumTimesOnShow),
        Date = max(Date), Dollars = sum(FinalScore)) %>%
    filter(MaxNumGames >= 10) %>%
    select(Name, MaxNumGames, Date, Dollars) %>%
    arrange(desc(MaxNumGames), desc(Date)) %>%
	as.data.frame()
```

The total winninges versus number of games played for players who won less than 10 games
(we use jitter to better visualize the data in each group):

```{r plot08, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
group_by(jeopardyData, PlayerId) %>%
    mutate(NumberWins = n() - 1) %>%
    summarise(Wins = first(NumberWins), Dollars = sum(FinalScore)) %>%
    filter(Wins > 0) %>%
    filter(Wins < 10) %>%
    ggplot(aes(x=Wins, y=Dollars)) +
        geom_jitter(position = position_jitter(width = .3)) +
        ggtitle("Total Amount Won by a Player versus Number of Games Played")
```

## Another Look at Winnings

```{r plot09, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
jeopardyData %>%
    group_by(PlayerId) %>%
    mutate(NumberWins = n() - 1) %>%
    summarise(Wins = first(NumberWins), Dollars = sum(FinalScore)) %>%
    filter(Wins > 0) %>%
    filter(Wins < 10) %>%
    ggplot(aes(x=Wins, y=Dollars)) +
        geom_boxplot(aes(group = Wins)) +
        ggtitle("Boxplots of Amount Won by a Player versus Number of Games Played")
```

## Winnings by  Gender

The gender of the
`r length(unique(jeopardyData$PlayerId))`
players:

```{r}
jeopardyData %>%
    filter(!is.na(Gender)) %>%
    group_by(PlayerId) %>%
    summarise(Gender1 = first(Gender)) %>%
    with(cbind(Count=table(Gender1),
        Percent=round(table(Gender1)/sum(table(Gender1)) * 100, 2)))
```

This is the gender of all the players
(that we were able to compute the gender for)
and is close to 50:50.
At least one third of all of these players will play in more than one game
(see earlier analysis of number of names played by gender).

Here we look at the gender distribution for each game.
There are 4 possible combinations for the 3 contestants:
FFF, FFM, FMM, and MMM.
Since the winner plays in the next game,
the same player can contribute to many gender tryads.

To compute:

```{r}
trigen <- function(gen) {
    if(any(is.na(gen))) {
      NA_character_
    } else {
        paste0(sort(gen), collapse = "")
    }
}
jeopardyData %>%
    group_by(Show) %>%
    summarise(ShowGender = trigen(Gender)) %>%
    with(cbind(Count=table(ShowGender),
        Percent=round(table(ShowGender)/sum(table(ShowGender)) * 100, 2))) %>%
	knitr::kable(padding=0)
```
A lot more games are played with two or more male contestants.

## Ken Jennings

The player with longest winning streak, 74 games
(the 75th game was a loss):

```{r}
jeopardyData %>%
    group_by(PlayerId) %>%
    mutate(MaxNumGames = max(NumTimesOnShow)) %>%
    filter(MaxNumGames == 75) %>%
    summarise(NumberWins = n() - 1) %>%
	as.data.frame()
```

Ken Jennings won
`r filter(jeopardyData,
    PlayerId == "Ken Jennings.software engineer.Salt Lake City.Utah",
        NumTimesOnShow < 75) %>%
    summarise(TotalWinnings = sum(FinalScore)) %>%
	pull() %>%
    format(scientific = FALSE, big.mark = ",")`
dollars in his 75 games.

## Ken Jennnings Final Game

The final scores in the game he lost:

```{r}
jeopardyData %>%
    group_by(Show) %>%
    filter("Ken Jennings.software engineer.Salt Lake City.Utah"
        %in% PlayerId, any(NumTimesOnShow == 75)) %>%
        ungroup() %>%
        select(Name, FinalScore) %>%
        arrange(desc(FinalScore)) %>%
		as.data.frame()
```

## Ken Jennings Winnings by Game

His winnings by game:

```{r plot10, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
filter(jeopardyData,
    PlayerId == "Ken Jennings.software engineer.Salt Lake City.Utah",
    NumTimesOnShow < 75) %>%
    ggplot(aes(x=Date, y=FinalScore)) + geom_point() +
    xlab("Date in 2004") +
    ggtitle("Ken Jennings Winnings by Date of Game")
```

## Ken Jennings Percent Correct

Jennings percent correct in each game:

```{r plot11, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
filter(jeopardyData,
    PlayerId == "Ken Jennings.software engineer.Salt Lake City.Utah") %>%
    mutate(PercentCorrect =
        (NumRight + DDRight) / (NumRight + NumWrong + DDWrong + DDRight) * 100) %>%
    with(summary(PercentCorrect))
```

## Ken Jennings Runaways

How often did Jennings win in a runaway?

```{r}
jeopardyData %>%
    group_by(Show) %>%
    filter("Ken Jennings.software engineer.Salt Lake City.Utah"
        %in% PlayerId, all(NumTimesOnShow < 75)) %>%
    summarise(Runaway = isRunaway(SecondRoundScore)) %>%
    with(c(NumberRunaways=sum(Runaway),
        PercentRunaways=round(sum(Runaway) / length(Runaway)  * 100, 2)))
```

## Players by State

Where do Jeopardy players come from?
While there are players from countries other than the United States,
the vast majority are from the U.S.

The number of players from each state;

```{r fig.width=8, fig.height=8}
NPlayers <- table(jeopardyData$State)
stateDF <- data.frame(State = names(NPlayers),
    NPlayers = c(NPlayers)) %>%
    filter(State %in% c(state.name, "D.C."))
ggplot(stateDF, aes(x=NPlayers,
        y=ordered(State, levels=State[order(NPlayers)]))) +
        geom_point() + xlab("Number of Players") + ylab("State") +
        ggtitle("Number of Players by State")
```

The state with more people tend to have more players on Jeopardy.
To make more appropriate comparisons, we need state population data.

State population data was downloaded from the U.S. Census
http://www.census.gov/popest/data/state/totals/2012/tables/NST-EST2012-01.csv.
The CSV file includes aggregation data at the top and population values
for multiple years.
We only extracted the data for the 50 states and the District of Columbia
and only used the 2010 values.

```{r}
# URL <- "http://www.census.gov/popest/data/state/totals/2012/tables/NST-EST2012-01.csv"
# censusTmpFile <- tempfile()
# download.file(URL, destfile=censusTmpFile)
# popPipe <- pipe(paste0("sed -n -e '/Alabama/,/Wyoming/p' ", censusTmpFile, " | cut -d'\"' --output-delimiter=\";\" -f1-2 | sed -e 's/\"//' -e 's/^.//' -e 's/,//g'"))
# statePop <- read.table(popPipe, sep=";", col.names=c("State", "Population"),
#     stringsAsFactors=FALSE)
statePop <- read.csv("pop_by_state_2017.csv", stringsAsFactors=FALSE)
statePop[statePop$State == "District of Columbia", "State"] <- "D.C."
indx51 <- match(stateDF$State, statePop$State)
stateDF$Population <- statePop[indx51, "Population"]
```

## Players vs State Population

The number of players versus the state population:

```{r plot12, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
ggplot(data=stateDF, aes(x=Population, y=NPlayers)) +
    geom_point() + xlab("State Population") +
    geom_text(data=subset(stateDF,
        State %in% c("New York", "D.C.", "Texas", "Florida")),
        aes(Population, NPlayers, label=State, vjust=1.3, hjust=0.2)) +
    geom_text(data=subset(stateDF,
        State %in% c("California")),
        aes(Population, NPlayers, label=State, vjust=1.3, hjust=0.8)) +
    ggtitle("Number of Players versus State Population")
```

As we noted earlier,
the bigger the state's population, the more players from that state.
Texas and Florida have the second and fourth largest population but
a low number of players and D.C. with a small population,
has a disproportionate number of players.

To see if players are equally likely to come from any U.S. state,
we want to compare the number of players from each state
to the state's population.
We add a standardized number of players per million people in a state to the
`stateDF` dataset:

## Normalized Players by State

```{r}
stateDF$PlayersPerMillion <- with(stateDF, NPlayers/Population * 1e6)
```

The number of players standardized by state's population (on log2 scale):

```{r plot13, fig.width=fwh, fig.height=fht, fig.show='hold', fig.align='center'}
ggplot(data=stateDF, aes(x=log2(PlayersPerMillion),
    y=ordered(State, levels=State[order(stateDF$PlayersPerMillion)]))) +
    geom_point() +
    xlab("Number of Players / 1 Million Population") + ylab("State") +
    ggtitle("Number of Players Standardized by State Population")
```

The East coast and West Coast is very well represented
after adjusting for population.
The Midwest is mostly in the middle and the South and plains states are
under represented.
