---
title: Analysis of Jeopardy Games
author: Stephen Kaluzny
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r echo=FALSE}
jeopardyDataRaw <- readRDS("jeopardyDataRaw.rds")
jeopardyData <- readRDS("jeopardyData.rds")
```

Jeopardy is a television quiz show where three contestants compete by
responding to general knowledge clues.
Their responses must be in the form of a question.
The daily syndicated version of the show has been on the air since 1984.

This paper is an analysis of data from the game show.

#### Game Description

The show has two rounds and a final.
Each round has 6 categories and within each category there are 5 clues.
The clues have increasing dollar amounts with a category.
If a player gives the correct answer to a clues, that dollar amount
is added to their score.
Incorrect answers result in subtraction of that dollar amount.

The final jeopardy clue ends the game.
The players are told the category and they can wager as much of their score.
The players have 30 seconds to write down their response once the final clue
is shown.
Players with a correct response get their wager added to their score,
incorrect responses result in subtraction of the wager amount.
The player with the highest score wins that amount in dollars and
plays in the next day's game against two new players.

Data from the Jeopardy games are are available on the Internet.
It includes information about the players:
names, home city, number of games played, winnings, etc.
This analysis does not look at the content of the game questions (answers)
which are also available on the site.

#### Purpose of This Analysis

This document is an example of data clean up, manipulation
and exploration using R.
The code for all of the data analysis is shown in the document.
This is an example of a reproducible analysis.
All code to recreate the analysis, from getting the data,
cleaning it, to creating graphics is included.

This was also an exercise in using the `dplyr` package
for data cleanup and manipulation and
the creating graphics with the `ggplot2` package.
We made extensive use of the new pipe operator (`%>%`)
that was introduced wtih the `dplyr` package.

Those key packages are loaded here.
```{r}
library(ggplot2, quietly= TRUE, warn.conflicts = FALSE)
library(dplyr, quietly= TRUE, warn.conflicts = FALSE)
```

## The Data
### Getting the Data
The data was scraped from
the http://www.j-archive.com site
using an R function that was modified from
https://scraperwiki.com/scrapers/jarchive_scraper_taiwanshaun/.
The function, `jscrapeDirect`, is available in the Appendix.
Only regular Jeopardy games were downloaded,
leaving out championships, teen, college, etc tournaments.
The `id` argument is a vector of show id's to scrape from the web site.

To avoid having to get the data multiple times while developing the code,
the `jscrapeDirect` function can be used in two steps.
First call it with `getDataOnly=TRUE` to get the raw scraped data.
Then call `jscrapeDirect` with the raw data as an input
(as the `data` argument)
into a second call to `jscrapeDirect` to process the raw data.

```{r eval = FALSE}
jeopardyDataRaw <- jscrapeDirect(id=1:5000)
```

### Cleaning the Data
The scraped data has 15 columns:
```{r}
names(jeopardyDataRaw)
```

From these columns additional variables were computed.
Variables were renamed to a consistent and descriptive naming scheme.

The resulting data needed to be cleaned up.
Changes made include values in the wrong column,
spelling error in names of cities and states,
New York city boroughs changed to New York,
"now living in ..." and "now stationed at ..."
changed to the listed location.
The cleaning was an iterative process,
looking at summaries of the data, finding more errors,
making corrections.
The clean up operations were all encapsulate in a single
function, `jcleanData` that could be repeatedly run on the
`jeopardyRawData` object.
The `jcleanData` function is listed in the appendix.

### Additional Variables
To identify individual players, a `PlayerId` was created by pasting together the `Name`, `Occupation`, `City`, and `State`:
```{r eval = FALSE}
jeopardyData$PlayerId <- with(jeopardyData,
    paste(Name, Occupation, City, State, sep='.'))
```
Analysis of `PlayerId` indicated that it uniquely identified players
e.g. show numbers and dates were grouped together.

A `Date` variable was extracted from information in the `Title` variable:

```{r eval = FALSE}
airedIndex <- regexpr(" aired ", jeopardyData$Title)
jeopardyData$Date <- as.Date(
    substr(jeopardyData$Title, airedIndex + 7, nchar(jeopardyData$Title)))
```

After November 23, 2001 the game doubled the dollar amounts for each answer.
A `GameMaxValue` variable was added to distinguish games before and after the change:

```{r eval = FALSE}
airedIndex <- regexpr(" aired ", jeopardyData$Title)
jeopardyData$Date <- as.Date(
    substr(jeopardyData$Title, airedIndex + 7, nchar(jeopardyData$Title)))
```

A variable, `IsWinner`, was added to identify the winner(s) in each game (tied final scores result in multiple winners):

```{r eval =FALSE}
isWinner <-
function (x)                                                                    
{
    r <- rank(x)
    if (all(r == 2)) {
        r <- c(4, 4, 4)
    }
    (r == 3) | (r == 2.5) | (r == 4)
}
jeopardyData <- group_by(jeopardyData, Show) %>%
    mutate(IsWinner = isWinner(FinalScore))
jeopardyData <- ungroup(jeopardyData)
```

### Gender
To analyze the results based on gender,
the gender of each player needed to determined.
We used the `gender` package from CRAN.
The `gender` function from that package takes a first name and returns
the proportion of males and females with that name
(along with other attributes).
We use a cutoff proportion of `0.75` - if the proportion for a gender was
above the cutoff, we use that gender.
If it was less, we set the gender to `NA`.
To save computing time, we only compute the gender for the unique first names and then applied the results to all players.

```{r eval = FALSE}
library(gender)
firstName <- with(jeopardyData, gsub("^(.*?)\\s.*", "\\1", Name))
# Estimate gender using the gender from the gender package
# Only do the unique firstNames
uFirstName <- unique(firstName)
# This takes 232.14 secs on boulder:
uGenderList <- gender(uFirstName)
uGender <- sapply(uGenderList, function(x) x$gender)
#
# Code "male" -> "M", "female" -> "F"
uGender <- ifelse(uGender == "male", "M", uGender)
uGender <- ifelse(uGender == "female", "F", uGender)
uPropMale <- sapply(uGenderList, function(x) x$proportion_male)
uPropFemale <- sapply(uGenderList, function(x) x$proportion_female)
#
# Only consider gender that has prop. greater than propCutoff,
# otherwise, declare it as NA
propCutoff <- 0.75
uGender <- ifelse((!is.na(uGender) & uGender == "M" &
    uPropMale < propCutoff), NA, uGender)
uGender <- ifelse((!is.na(uGender) & uGender == "F" &
    uPropFemale < propCutoff), NA, uGender)
mFirstName <- match(firstName, uFirstName)
jeopardyData$Gender <- uGender[mFirstName]
```
We were able to predict gender for
`r with(jeopardyData, sum(!duplicated(PlayerId) & !is.na(jeopardyData$Gender)))`
players which is
`r round(with(jeopardyData, sum(!duplicated(PlayerId) & !is.na(jeopardyData$Gender)) / sum(!duplicated(PlayerId))) *100, 2)`
percent of the players.

The final dataset had `r ncol(jeopardyData)`  variables:
```{r}
names(jeopardyData)
```
## Analysis

The final dataset for analysis has
`r nrow(jeopardyData)` rows and `r ncol(jeopardyData)` columns.
Each row represents a single player in a game,
each games consists of 3 rows
i,e, we have
`r nrow(jeopardyData)` player observations and
`r nrow(jeopardyData) / 3` games.
There are `r length(unique(jeopardyData$PlayerId))` different players.
The games were played from
`r min(jeopardyData$Date)` to `r max(jeopardyData$Date)`.

### The Games

#### Answers

There are 6 categories x 5 clues per category x 2 rounds = 60 clues
in each game.

Due to time limitations, the players may not get to all of the clues.
The number of clues attempted in a game by a single player is
the sum of the number correct (`NumRight`), the number wrong (`NumWrong`),
the number of daily doubles correct (`DDRight`)
and the number of daily doubles wrong (`DDWrong`).

```{r}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight) %>%
    with(summary(NumAttempted))
```

At least one player only attempted 2 clues and while another dominated
the game by trying 51.

A graphical view of the number of attempted clues:

```{r fig.width=8, fig.height=7}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight) %>%
    ggplot(aes(x=factor(NumAttempted))) + 
        geom_bar() + xlab("Number Clues Attempted by a Player") +
        ggtitle("Distribution of Number of Clues Attempted by a Player")
```

The same graph, by gender:

```{r fig.width=8, fig.height=7}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight) %>%
    filter(!is.na(Gender)) %>%
    ggplot(aes(x=factor(NumAttempted))) + 
        geom_bar() + xlab("Number Clues Attempted by a Player") +
        facet_grid(Gender ~ .) +
        ggtitle("Distribution of Clues Attempted by Gender")

```

Men tend to attempt more questions then women
and the range of values is wider for men including
some that attempt over 50 of the possible questions.

Note that the sum of `NumAttempted` across all three players in a game
is not the same as the total number of clues tried in that game
since after an incorrect response by a players,
another play can try to answer that same question.

The percentage of clues answers correctly by each player:

```{r}
jeopardyData %>%
    mutate(PercentCorrect =
        (NumRight + DDRight) / (NumRight + NumWrong + DDWrong + DDRight) * 100) %>%
    with(summary(PercentCorrect))
```
A density plot of the percent correct answers:

```{r fig.width=8, fig.height=7}
jeopardyData %>%
    mutate(PercentCorrect =
        (NumRight + DDRight) /
            (NumRight + NumWrong + DDWrong + DDRight) * 100) %>%
    ggplot(aes(x=PercentCorrect)) + 
        geom_density()
```

The most common percent correct values are between 82 and 90,
then it drops off.
There is a bump at 100 percent.


Here is the distribution of attempted clues for those who always 
answer correctly compared to other players. We use a density plot
instead of counts since the total numbers in each group are vastly
different.

```{r fig.width=8, fig.height=7}
jeopardyData %>%
    mutate(NumAttempted = NumRight + NumWrong + DDWrong + DDRight,
        PercentCorrect = (NumRight + DDRight) /
            (NumRight + NumWrong + DDWrong + DDRight) * 100,
        AlwaysRight = (PercentCorrect > 99)) %>%
    ggplot(aes(x = NumAttempted, color=AlwaysRight)) +
        geom_density() +
        xlab("Number of Clues Attempted") +
        ggtitle("Number Attempted for Perfect Player vs Other Players")
```

Most players who get all questions correct tend not to attempt as many questions as other players.

Here is the density of percent correct answers by gender:

```{r fig.width=8, fig.height=7}
jeopardyData %>%
    filter(!is.na(Gender)) %>%
    mutate(PercentCorrect =
        (NumRight + DDRight) /
            (NumRight + NumWrong + DDWrong + DDRight) * 100) %>%
    ggplot(aes(x=PercentCorrect)) + 
        geom_density() + facet_grid(Gender ~ .) +
        ggtitle("Density of Percent Correct by Gender")
```

We have the spike at 100 percent for both female and male players.

#### Daily Doubles

There are 3 daily doubles in each game, one in the Jeopardy Round and
two in the Double Jeopardy round.
The player who gets the daily double questions 
can wager as much of their current score on the question.

The number of daily doubles found by players in a game:

```{r}
jeopardyData %>%
    mutate(DDCount = DDWrong + DDRight) %>%
    with(cbind(NumDD = 0:3, DDCount = table(DDCount),
       Percent =  round(table(DDCount) /  sum(table(DDCount)) * 100, 2))) 
```
The percentage of all daily doubles that are answered correctly is
`r 
jeopardyData %>%
    filter(DDWrong + DDRight > 0) %>%
    summarise(PercentDDRight = round(sum(DDRight) / sum(DDWrong + DDRight) * 100, 2))
` percent.

#### Number of Final Jeopardy Players

If a player has a score of 0 or less at the end of the double jeopardy round,
that player is eliminated from the final jeopardy round
The number of players in Final Jeopardy:

```{r}
NFinal <- jeopardyData %>%
    group_by(Show) %>%
    mutate(NumberInFinal = sum(SecondRoundScore > 0)) %>%
    summarise(NumberInFinal = first(NumberInFinal)) %>%
    with(table(NumberInFinal))
cbind(NumberOfPlayers=1:3, Count = NFinal,
    Percentage = round(NFinal / sum(NFinal) * 100, 2))
```

The show with only one player left for Final Jeopardy:
```{r}
jeopardyData %>%
    group_by(Show) %>%
    mutate(NumberInFinal = sum(SecondRoundScore > 0)) %>%
    filter(NumberInFinal == 1) %>%
    select(Date, Name, SecondRoundScore) %>%
    as.data.frame()
```

#### Runaways

A runaway game is when one player cannot be beat in Final Jeopardy.
This will occur when one player has at least twice the score of each of the
other players at the end of the Double Jeopardy round.

```{r}
isRunaway <- function(score) {
    score <- sort(score)
    all(score[3] >= 2*score[-3])
}
jeopardyData %>%
    group_by(Show) %>%
    summarise(Runaway = isRunaway(SecondRoundScore)) %>%
    with(c(NumberRunaways=sum(Runaway),
        PercentRunaways=round(sum(Runaway) / length(Runaway) * 100, 2)))
```

#### Ties
How often does the game end with multiple winners (i.e. a tie)?

```{r}
jeopardyData %>%
    group_by(Show) %>%
    mutate(NumWinners = sum(IsWinner)) %>%
    summarise(NumWinners = first(NumWinners)) %>%
    with(cbind(Winners = 1:3, Count = table(NumWinners),
        Percent = round(table(NumWinners) / sum(NumWinners) * 100, 2)))
```

Most games have only one winner.
In two games, all three people had the same final score.

#### Comebacks

How often does the player with the lowest score
at the end of the double jeopardy round, win the game?

```{r}
jeopardyData %>%
    group_by(Show) %>%
    mutate(IsComeback = (SecondRoundScore == min(SecondRoundScore))) %>%
    filter(IsWinner) %>%
    with(c(Count = sum(IsComeback),
        Percent = round(sum(IsComeback) / length(IsComeback) * 100, 2)))
```

#### Final Jeopardy Stumpers

How often do all the players miss the Final Jeopardy questions?
The data does not include any specific information about the Final Jeopardy question
or wager.
If the `FinalScore` is larger than the `SecondRoundScore`, we know the player answered
Final Jeopardy correctly. Similarly, if `FinalScore` is less than `SecondRoundScore`,
their response for Final Jeopardy was incorrect.
If the `FinalScore` is the same as the `SecondRoundScore`, we only know that the
player did not wager anything in Final Jeopardy, their response could have been correct or wrong.
So we only count those who made a wager in Final Jeopardy.

```{r}
isStumper <- function(score2, score3) {
    all(score3 < score2)
}
jeopardyData %>%
    group_by(Show) %>%
    filter(SecondRoundScore != FinalScore) %>%
    summarise(IsStumper = isStumper(SecondRoundScore, FinalScore)) %>%
    with(c(Count = sum(IsStumper), Games = length(IsStumper),
        Percent = round(sum(IsStumper) / length(IsStumper) * 100, 2)))
```

#### Dollar Winnings in a Game

When looking at the dollar winnings for a game we need to look separately
at games before the dollar values doubled and after.

```{r}
jeopardyData %>%
    filter(IsWinner) %>%
    group_by(GameMaxValue) %>%
    summarize(AveWinnings=mean(FinalScore),
        MedianWinnings=median(FinalScore),
        MinWinnings=min(FinalScore),
        MaxWinnings=max(FinalScore)) %>%
    as.data.frame()
```

The mean and median winnings are both about doubled
when the dollar amounts doubled.

The shows where the winning score was 0:

```{r}
jeopardyData %>%
    filter(IsWinner, FinalScore == 0) %>%
    select(Date, Name, FinalScore) %>%
    as.data.frame()
```

In one game, all 3 players won with 0
and in another game, two players tied with a score of 0.
Actually, none of the players were considered winners
since none of these players played in the next day's game.

A visual view of the distribution of the winnings:

```{r fig.width=8, fig.height=7}
filter(jeopardyData, IsWinner) %>% 
    ggplot(aes(x=FinalScore)) +
    geom_histogram(binwidth = 2000) +
    facet_grid(GameMaxValue ~ .) +
    ggtitle("Distribution of Game Winnings")
```

There are
`filter(jeopardyData, GameMaxValue == "Max2000") %>% nrow()`
with GameMaxValue of Max2000
compared to only
`filter(jeopardyData, GameMaxValue == "Max1000") %>% nrow()`
for Max1000.
It is no surprising that the spread of the FinalScore values
is larger for the Max2000 group.

### The Players

#### Number of Games Played

The `NumTimesOnShow` variable increases each time the player is on the show.
We want to look at the maximum value of `NumTimeOnShow` for each player:

```{r}
jeopardyData %>%
    group_by(PlayerId) %>%
    mutate(MaxNumGames = max(NumTimesOnShow)) %>%
    with(summary(MaxNumGames))
```

Half the players only are on one show.
Which is not surprising since two players (usually) do not return.
One player was on 75 times.
We will investigate that player later.

Number of games played by gender:

```{r}
jeopardyData %>%
    filter(!is.na(Gender)) %>%
    group_by(PlayerId) %>%
    mutate(MaxNumGames = max(NumTimesOnShow)) %>%
    group_by(Gender) %>%
    summarise(Min=min(MaxNumGames), Median=median(MaxNumGames),
        Mean=mean(MaxNumGames), Max=max(MaxNumGames))
```

Looking at the players who won the most games
(the top 6, the Date is the date of ther final game):

```{r}
jeopardyData %>%
    group_by(PlayerId) %>%
    summarise(Name = first(Name), MaxNumGames = max(NumTimesOnShow),
        Date = max(Date)) %>%
    top_n(6, MaxNumGames) %>%
    select(Name, MaxNumGames, Date) %>%
    arrange(desc(MaxNumGames), desc(Date))
 ```

The distribution of the number of games by a player,
leaving out the outliers (players who played more than 10 games)
as well as players who never won (i.e. only played one game):

```{r fig.width=8, fig.height=7}
jeopardyData %>%
    group_by(PlayerId) %>%
    summarise(MaxNumGames = max(NumTimesOnShow)) %>%
    filter(MaxNumGames > 1 & MaxNumGames < 11) %>%
    ggplot(aes(x=factor(MaxNumGames))) +
        geom_bar(binwidth=0.10) + xlab("NumGamesPlayed") +
        coord_flip() +
        ggtitle("Counts of Number of Games Played by a Player")
```

Most winners only win once and winning more than 5 games is rare.

#### Total Winnings Per Player 

Here we look at the total winnings for player over all the
games he or she played.
The number of wins is one less than the number of times on the show.
We use jitter to better visualize the data in each group:

```{r fig.width=8, fig.height=7}
group_by(jeopardyData, PlayerId) %>%
    mutate(NumberWins = n() - 1) %>%
    summarise(Wins = first(NumberWins), Dollars = sum(FinalScore)) %>%
    filter(Wins > 0) %>%
    filter( Wins < 19) %>%
    ggplot(aes(x=Wins, y=Dollars)) +
        geom_jitter(position = position_jitter(width = .3)) +
        ggtitle("Total Amount Won by a Player versus Number of Games Played")
```

Alternatively, we could use boxplots for the same data:

```{r fig.width=8, fig.height=7}
jeopardyData %>%
    group_by(PlayerId) %>%
    mutate(NumberWins = n() - 1) %>%
    summarise(Wins = first(NumberWins), Dollars = sum(FinalScore)) %>%
    filter(Wins > 0) %>%
    filter(Wins < 19) %>%
    ggplot(aes(x=Wins, y=Dollars)) +
        geom_boxplot(aes(group = Wins)) +
        ggtitle("Boxplots of Amount Won by a Player versus Number of Games Played")
```

#### Gender

The gender of the
`r length(unique(jeopardyData$PlayerId))`
players:

```{r}
jeopardyData %>%
    filter(!is.na(Gender)) %>%
    group_by(PlayerId) %>%
    summarise(Gender1 = first(Gender)) %>%
    with(cbind(Count=table(Gender1),
        Percent=round(table(Gender1)/sum(table(Gender1)) * 100, 2)))
```

This is the gender of all the players
(that we were able to compute the gender for)
and is close to 50:50.
At least one third of all of these players will play in more than one game
(see earlier analysis of number of names played by gender).

Here we look at the gender distribution for each game.
There are 4 possible combinations for the 3 contestants:
FFF, FFM, FMM, and MMM.
Since the winner plays in the next game,
the same player can contribute to many gender tryads.

To compute:

```{r}
trigen <- function(gen) {
    if(any(is.na(gen))) {
      NA_character_
    } else {
        paste0(sort(gen), collapse = "")
    }
}
jeopardyData %>%
    group_by(Show) %>%
    summarise(ShowGender = trigen(Gender)) %>%
    with(cbind(Count=table(ShowGender),
        Percent=round(table(ShowGender)/sum(table(ShowGender)) * 100, 2)))
```
A lot more games are played with two or more male contestants.

### Ken Jennings

The player with longest winning streak, 74 games
(the 75th game was a loss):

```{r}
jeopardyData %>%
    group_by(PlayerId) %>%
    mutate(MaxNumGames = max(NumTimesOnShow)) %>%
    filter(MaxNumGames == 75) %>%
    summarise(NumberWins = n() - 1)
```

Ken Jennings won
`r filter(jeopardyData,
    PlayerId == "Ken Jennings.software engineer.Salt Lake City.Utah",
        NumTimesOnShow < 75) %>%
    summarise(TotalWinnings = sum(FinalScore)) %>%
    format(scientific = FALSE, big.mark = ",")`
dollars in his 75 games.

The final scores in the game he lost:

```{r}
jeopardyData %>%
    group_by(Show) %>%
    filter("Ken Jennings.software engineer.Salt Lake City.Utah"
        %in% PlayerId, any(NumTimesOnShow == 75)) %>%
        ungroup() %>%
        select(Name, FinalScore) %>%
        arrange(desc(FinalScore))
```

His winnings by game:

```{r fig.width=8, fig.height=7}
filter(jeopardyData,
    PlayerId == "Ken Jennings.software engineer.Salt Lake City.Utah",
    NumTimesOnShow < 75) %>%
    ggplot(aes(x=Date, y=FinalScore)) + geom_point() +
    xlab("Date in 2004") +
    ggtitle("Ken Jennings Winnings by Date of Game")
```

Jennings percent correct in each game:

```{r fig.width=8, fig.height=7}
filter(jeopardyData,
    PlayerId == "Ken Jennings.software engineer.Salt Lake City.Utah") %>%
    mutate(PercentCorrect =
        (NumRight + DDRight) / (NumRight + NumWrong + DDWrong + DDRight) * 100) %>%
    with(summary(PercentCorrect))
```
    
How often did Jennings win in a runaway?

```{r}
jeopardyData %>%
    group_by(Show) %>%
    filter("Ken Jennings.software engineer.Salt Lake City.Utah"
        %in% PlayerId, all(NumTimesOnShow < 75)) %>%
    summarise(Runaway = isRunaway(SecondRoundScore)) %>%
    with(c(NumberRunaways=sum(Runaway),
        PercentRunaways=round(sum(Runaway) / length(Runaway)  * 100, 2)))
```

## Players by State

Where do Jeopardy players come from?
While there are players from countries other than the United States,
the vast majority are from the U.S.

The number of players from each state;

```{r fig.width=8, fig.height=8}
NPlayers <- table(jeopardyData$State)
stateDF <- data.frame(State = names(NPlayers),
    NPlayers = c(NPlayers)) %>%
    filter(State %in% c(state.name, "D.C."))
ggplot(stateDF, aes(x=NPlayers,
        y=ordered(State, levels=State[order(NPlayers)]))) +
        geom_point() + xlab("Number of Players") + ylab("State") +
        ggtitle("Number of Players by State")
```

The state with more people tend to have more players on Jeopardy.
To view more formally, we need state population data.

State population data was downloaded from the U.S. Census
http://www.census.gov/popest/data/state/totals/2012/tables/NST-EST2012-01.csv.
The CSV file includes aggregation data at the top and population values
for multiple years.
We only extracted the data for the 50 states and the District of Columbia
and only used the 2010 values.

```{r cache = TRUE}
URL <- "http://www.census.gov/popest/data/state/totals/2012/tables/NST-EST2012-01.csv"
censusTmpFile <- tempfile()
download.file(URL, destfile=censusTmpFile)
popPipe <- pipe(paste0("sed -n -e '/Alabama/,/Wyoming/p' ", censusTmpFile, " | cut -d'\"' --output-delimiter=\";\" -f1-2 | sed -e 's/\"//' -e 's/^.//' -e 's/,//g'"))
statePop <- read.table(popPipe, sep=";", col.names=c("State", "Population"),
    stringsAsFactors=FALSE)
statePop[statePop$State == "District of Columbia", "State"] <- "D.C."
indx51 <- match(stateDF$State, statePop$State)
stateDF$Population <- statePop[indx51, "Population"]
```

The number of players versus the state population:

```{r fig.width=8, fig.height=7}
ggplot(data=stateDF, aes(x=Population, y=NPlayers)) + 
    geom_point() + xlab("State Population") +
    geom_text(data=subset(stateDF,
        State %in% c("New York", "D.C.", "Texas", "Florida")),
        aes(Population, NPlayers, label=State, vjust=1.3, hjust=0.2)) +
    geom_text(data=subset(stateDF,
        State %in% c("California")),
        aes(Population, NPlayers, label=State, vjust=1.3, hjust=0.8)) +
    ggtitle("Number of Players versus State Population")
```

As we noted earlier,
the bigger the state's population, the more players from that state.
Texas and Florida have the second and fourth largest population but
a low number of players and D.C. with a small population,
has a disproportionate number of players.

To see if players are equally likely to come from any U.S. state,
we want to compare the number of players from each state
to the state's population.
We add a standardized number of players per million people in a state to the
stateDF dataset:

```{r}
stateDF$PlayersPerMillion <- with(stateDF, NPlayers/Population * 1e6)
```

The number of players standardized by state's population (on log2 scale):

```{r fig.width=8, fig.height=8}
ggplot(data=stateDF, aes(x=log2(PlayersPerMillion),
    y=ordered(State, levels=State[order(stateDF$PlayersPerMillion)]))) +
    geom_point() +
    xlab("Number of Players / 1 Million Population") + ylab("State") +
    ggtitle("Number of Players Standardized by State Population")
```

The East coast is very well represented after adjusting for population.
The Midwest is mostly in the middle and the South and plains states are
are under represented.

The same plot with `D.C.` left out:

```{r fig.width=8, fig.height=8}
stateDF %>%
  filter(State != "D.C.") %>%
  ggplot(aes(x=log2(PlayersPerMillion),
    y=ordered(State, levels=State[order(PlayersPerMillion)]))) +
    geom_point()+ xlab("Number of Players / 1 Million Population") + ylab("State") +
    ggtitle("Number of Players Standardized by State Population Omitting D.C")
```

## Appendix

The web scraper function:
```{r eval=FALSE}
```
The data cleaning function:
```{r eval=FALSE}
```
